Floral resource choice by bumble bee species in the Berchtesgadener Alps

### 2020-12-02 analyses/niche_relations_v6.Rmd
update lab deskttop

## Define functions
```{r, echo = FALSE}
# Fortify envfit() object fitted to BB tongue length data
envfit_fort <- function(x){
  fortify(x) %>%
  select(Score = Type, Label, MDS1, MDS2) %>%
  mutate(Label = case_when(Label == "pbl.w" ~ "T.len"),
         Score = case_when(Score == "Vector" ~ "T.len"))
}

@@

# Plot PCoA ordination with arrow for tongue length envfit()
pcoa_plot <- function(x, y, min_load = 0.25) {
  ggplot(filter(x, (abs(MDS1) > min_load | abs(MDS2) > min_load) | (Score %in% c("BB.sp", "T.len")))) + # to avoid overplotting, we can filter to include only floral species > loading threshold
    geom_point(aes(MDS1, MDS2, color = Score)) +
    geom_text_repel(aes(MDS1, MDS2, color = Score, label = Label), key_glyph = "rect") +
    geom_segment(data = y, 
               aes(x = 0, y = 0, xend = MDS1, yend = MDS2, colour = Label),
               arrow = arrow(), size = 2, alpha = 0.5, key_glyph = "rect") +
    theme_light() +
    coord_fixed() # Gavin Simpson says this is **really** important
}

@@

  x %>%
    filter(caste != "male") %>%
    filter(!bb.sp %in% c("psit", "gers", "humi")) %>% # omit parasitic species, gers because it's a specialist, humi because it's so rare
    group_by(..., bb.sp, !!criterion) %>% # pooling all sites and years
    summarize(freq = n()) %>% # get interaction frequency for each BB-FL pair
    group_by(..., bb.sp) %>%
    mutate(prop = freq/sum(freq)) %>% # get proportional interaction frequency for each BB-FL pair
    dplyr::select(-freq) %>% # drop freq variable
    group_by(...) %>%
    nest() %>% # nest into list-column
    mutate(comm1 = map(data, # spread into the kind of array vegan likes
                       function(x) pivot_wider(x, names_from = !!criterion, values_from = prop) %>%
                         replace(is.na(.), 0)
                       ), 
           comm2 = map(comm1, # column to rowname
                       function(x) column_to_rownames(x, var = "bb.sp")
                       ),
           pcoa = map(comm2, # call PCoA
                      function(x) capscale(x ~ 1, distance = "horn")
                      ),
           traits = map(comm1, # join BB traits
                        function(x) x %>% 
                          left_join(y) %>% # y is the trait table, an argument in the main function call
                          column_to_rownames(var = "bb.sp") %>% 
                          select(pbl.w) # pbl.w is the variable I'm interested in
                        ),
           adonis = map2(comm2, traits, # run adonis permuatation test for distance ~ tongue class
                         function(x, y) adonis(x ~ pbl.w, data = y, method = "horn")
                         ),
           envfit = map2(pcoa, traits, function(x, y) envfit(x, y)),
           envfit.fort = map(envfit, envfit_fort),
           pcoa.fort = map2(pcoa, envfit.fort, pcoa_fort),
           pcoa.plot = map2(pcoa.fort, envfit.fort, pcoa_plot), 
           envfit.r2 = map(envfit, function(x) x$vectors$r),
           envfit.p = map(envfit, function(x) x$vectors$pvals),
           adonis.r2 = map(adonis, # extract adonis r2
                           function(x) x$aov.tab$R2[[1]]
                           ),
           adonis.p = map(adonis, # extract adonis p-value
                          function(x) x$aov.tab$'Pr(>F)'[[1]]
                          ),
           ordisurf = map2(pcoa, traits, function(x, y) ordisurf(x ~ pbl.w, data = y, k = 8)),
           ordisurf.dev_exp = map(ordisurf, function(x) summary(x)$dev.expl),
           ordisurf.p = map(ordisurf, function(x) summary(x)$s.pv),
           MDS1 = map(pcoa, # extract variance explained by MDS1
                      function(x) x$CA$eig[[1]] / sum(x$CA$eig)
                      ), 
           MDS2 = map(pcoa, # extract variance explained by MDS2
                      function(x) x$CA$eig[[2]] / sum(x$CA$eig)
                      )) %>%
    unnest(cols = c(ordisurf.dev_exp, ordisurf.p, envfit.r2, envfit.p, adonis.r2, adonis.p, MDS1, MDS2)) %>% # these are lists of length = 1, so unnesting them simply displays the value
    select(..., comm2, traits, pcoa, ordisurf, ordisurf.dev_exp, ordisurf.p, envfit, envfit.fort, pcoa.fort, pcoa.plot, envfit.r2, envfit.p, adonis.r2, adonis.p, MDS1, MDS2)
}

ordi1 <- ordisurf(niche.tax$pcoa[[1]] ~ pbl.w, data = niche.tax$traits[[1]])

@@

## 1.1 Ordination by species with post-hoc fitting of tongue length  
### Script
```{r message=FALSE, warning=FALSE, echo = FALSE}
# By plant.sp
niche.tax <- niche_analysis(net, bb_traits, criterion = plant.sp.abb)
niche.tax.plot <- niche.tax$pcoa.plot[[1]]

@@

# Playing with ordisurf as alternative to envfit()
library(ggordiplots)

## Based on https://oliviarata.wordpress.com/2014/07/17/ordinations-in-ggplot2-v2-ordisurf/
# Get ordination object
ordi1 <- niche.tax$ordisurf[[1]] 
ordi1.grid <- ordi1$grid #extracts the ordisurf object
ordi1.mite <- expand.grid(x = ordi1.grid$x, y = ordi1.grid$y) #get x and ys
ordi1.mite$z <- as.vector(ordi1.grid$z) #unravel the matrix for the z scores
ordi1.mite.na <- data.frame(na.omit(ordi1.mite)) %>% #gets rid of the nas
  as_tibble() %>%
  rename(MDS1 = x, MDS2 = y)

# Get scores
scores <- fortify(niche.tax$pcoa[[1]], axes = 1:2) %>%
  mutate(Score = case_when(
    Score == "species" ~ "FL.sp",
    Score == "sites" ~ "BB.sp")
    ) %>%
  as_tibble()

# Plot
dat <- filter(scores, (abs(MDS1) > 0.25 | abs(MDS2) > 0.25) | Score == "BB.sp")

ggplot(dat, aes(MDS1, MDS2, pch = Score)) +
  geom_contour(data = ordi1.mite.na, aes(MDS1, MDS2, z = z, color = stat(level)), size = 1.5, alpha = 0.75, inherit.aes = FALSE) +
  geom_point(size = 2) +
  geom_text_repel(aes(MDS1, MDS2, label = Label), key_glyph = "rect") +
  theme_light() +
  scale_color_viridis_c(name = "Tongue length") +
  scale_shape_manual(values = c(16, 1), name = "Level") +
  coord_fixed() # Gavin Simpson says this is **really** important
```

>>>>

## Define functions
```{r, echo = FALSE}
# Plot ordisurf
## Based on https://oliviarata.wordpress.com/2014/07/17/ordinations-in-ggplot2-v2-ordisurf/
surfplot <- function(x, y, min_score = 0.25) {

  grid <- x$grid # extracts the ordisurf object
  grid_exp <- expand.grid(x = grid$x, y = grid$y) # get x and ys
  grid_exp$z <- as.vector(grid$z) # unravel the matrix for the z scores

  contours <- data.frame(na.omit(grid_exp)) %>% # gets rid of the nas
    as_tibble() %>%
    rename(MDS1 = x, MDS2 = y)

  scores <- y %>% 
    filter(Score != "T.len") %>%
    filter((abs(MDS1) > 0.25 | abs(MDS2) > 0.25) | Score == "BB.sp")

  ggplot(scores, aes(MDS1, MDS2, pch = Score)) +
    geom_contour(data = contours, aes(MDS1, MDS2, z = z, color = stat(level)), size = 1.5, alpha = 0.75, inherit.aes = FALSE) +
    geom_point(size = 2) +
    geom_text_repel(aes(MDS1, MDS2, label = Label), key_glyph = "rect") +
    theme_light() +
    scale_color_viridis_c(name = "Tongue length") +
    scale_shape_manual(values = c(16, 1), name = "Level") +
    coord_fixed() # Gavin Simpson says this is **really** important
}

# Fortify envfit() object fitted to BB tongue length data
envfit_fort <- function(x){
  fortify(x) %>%
  select(Score = Type, Label, MDS1, MDS2) %>%
  mutate(Label = case_when(Label == "pbl.w" ~ "T.len"),
         Score = case_when(Score == "Vector" ~ "T.len"))
}

@@

# Plot PCoA ordination with arrow for tongue length envfit()
pcoa_plot <- function(x, y, min_score = 0.25) {
  ggplot(filter(x, (abs(MDS1) > min_score | abs(MDS2) > min_score) | (Score %in% c("BB.sp", "T.len")))) + # to avoid overplotting, we can filter to include only floral species > loading threshold
    geom_point(aes(MDS1, MDS2, color = Score)) +
    geom_text_repel(aes(MDS1, MDS2, color = Score, label = Label), key_glyph = "rect") +
    geom_segment(data = y, 
               aes(x = 0, y = 0, xend = MDS1, yend = MDS2, colour = Label),
               arrow = arrow(), size = 2, alpha = 0.5, key_glyph = "rect") +
    theme_light() +
    coord_fixed() # Gavin Simpson says this is **really** important
}

@@

  x %>%
    filter(caste != "male") %>%
    filter(!bb.sp %in% c("psit", "gers", "humi")) %>% # omit parasitic species, gers because it's a specialist, humi because it's so rare
    group_by(..., bb.sp, !!criterion) %>% # pooling all sites and years
    summarize(freq = n()) %>% # get interaction frequency for each BB-FL pair
    group_by(..., bb.sp) %>%
    mutate(prop = freq/sum(freq)) %>% # get proportional interaction frequency for each BB-FL pair
    dplyr::select(-freq) %>% # drop freq variable
    group_by(...) %>%
    nest() %>% # nest into list-column
    mutate(comm1 = map(data, # spread into the kind of array vegan likes
                       function(x) pivot_wider(x, names_from = !!criterion, values_from = prop) %>%
                         replace(is.na(.), 0)
                       ), 
           comm2 = map(comm1, # column to rowname
                       function(x) column_to_rownames(x, var = "bb.sp")
                       ),
           pcoa = map(comm2, # call PCoA
                      function(x) capscale(x ~ 1, distance = "horn")
                      ),
           traits = map(comm1, # join BB traits
                        function(x) x %>% 
                          left_join(y) %>% # y is the trait table, an argument in the main function call
                          column_to_rownames(var = "bb.sp") %>% 
                          select(pbl.w) # pbl.w is the variable I'm interested in
                        ),
           adonis = map2(comm2, traits, # run adonis permuatation test for distance ~ tongue class
                         function(x, y) adonis(x ~ pbl.w, data = y, method = "horn")
                         ),
           envfit = map2(pcoa, traits, function(x, y) envfit(x, y)),
           envfit.fort = map(envfit, envfit_fort),
           pcoa.fort = map2(pcoa, envfit.fort, pcoa_fort),
           pcoa.plot = map2(pcoa.fort, envfit.fort, pcoa_plot), 
           envfit.r2 = map(envfit, function(x) x$vectors$r),
           envfit.p = map(envfit, function(x) x$vectors$pvals),
           adonis.r2 = map(adonis, # extract adonis r2
                           function(x) x$aov.tab$R2[[1]]
                           ),
           adonis.p = map(adonis, # extract adonis p-value
                          function(x) x$aov.tab$'Pr(>F)'[[1]]
                          ),
           ordisurf = map2(pcoa, traits, function(x, y) ordisurf(x ~ pbl.w, data = y, k = 8)),
           ordisurf.dev_exp = map(ordisurf, function(x) summary(x)$dev.expl),
           ordisurf.p = map(ordisurf, function(x) summary(x)$s.pv),
           ordisurf.plot = map2(ordisurf, pcoa.fort, surfplot),
           MDS1 = map(pcoa, # extract variance explained by MDS1
                      function(x) x$CA$eig[[1]] / sum(x$CA$eig)
                      ), 
           MDS2 = map(pcoa, # extract variance explained by MDS2
                      function(x) x$CA$eig[[2]] / sum(x$CA$eig)
                      )) %>%
    unnest(cols = c(ordisurf.dev_exp, ordisurf.p, envfit.r2, envfit.p, adonis.r2, adonis.p, MDS1, MDS2)) %>% # these are lists of length = 1, so unnesting them simply displays the value
    select(..., comm2, traits, pcoa, ordisurf, ordisurf.dev_exp, ordisurf.p, ordisurf.plot, envfit, envfit.fort, pcoa.fort, pcoa.plot, envfit.r2, envfit.p, adonis.r2, adonis.p, MDS1, MDS2)
}

@@

## 1.1 Ordination by species with post-hoc fitting of tongue length  
### Script
```{r message=FALSE, warning=FALSE, echo = FALSE}
# By plant.sp
niche.tax <- niche_analysis(net, bb_traits, criterion = plant.sp.abb)
niche.tax.plot <- niche.tax$pcoa.plot[[1]]
niche.tax.surfplot <- niche.tax$ordisurf.plot
nich.tax.baseplot <- plot(niche.tax$ordisurf[[1]])
niche.tax$ordisurf.plot
plot(niche.tax$ordisurf[[1]])

@@
